---
title: "Effectivness of an Exercise using Predictive Analytics"
output:
  pdf_document: default
  html_document: default
date: "2024-12-11"
---


## Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. The following steps would be followed:

1. Data Pre-processing and Cleaning
2. Predictive Model Selection 
3. Validating Model and creating plots
4. Testing Model 
5. Predicting Results



## 1. Data Pre-processing and Cleaning
We load the training and testing datasets from the online sources. Next, we split the training dataset further into separate subsets for training, validation, and testing purposes.



```{r}
library(caret)
library(readxl)
acc_data_train <- read.csv("C:\\Users\\dhruv\\Downloads\\pml-training.csv")
acc_data_test <- read.csv("C:\\Users\\dhruv\\Downloads\\pml-testing.csv")
```

The data consists of many columns with near Zero variances , hence to improve our model we will get rid of these variables 


```{r}
NZ <- nearZeroVar(acc_data_train)
NZ1 <- nearZeroVar(acc_data_test)
acc_data_train <- acc_data_train[,-NZ]
acc_data_test <- acc_data_test[,-NZ]
```

Removing Variables with NA's

```{r}
label <- apply(acc_data_train, 2, function(x) mean(is.na(x))) > 0.95
label1 <- apply(acc_data_test, 2, function(x) mean(is.na(x))) > 0.95
acc_data_train <- acc_data_train[, -which(label, label == FALSE)]
acc_data_test <- acc_data_test[, -which(label1, label == FALSE)]
```

## 2. Predictive Model Selection : Random Forest
Random Forest is an ensemble learning method used for classification, regression, and other tasks. It works by constructing multiple decision trees during training and combining their outputs (e.g., majority voting for classification or averaging for regression) to improve accuracy and reduce overfitting. The algorithm introduces randomness by selecting a random subset of features for splitting at each tree node, making it robust to noise and capable of handling large datasets with high dimensionality.

```{r}
modFit <- train(classe ~ roll_belt, data = acc_data_train , method = "rf")
modFit$finalModel
```

## 3. Validating Model and creating plots

Creating a Validation Set within the training set to validate results from the predictive model built
```{r}
inTrain <- createDataPartition(y= acc_data_train$classe ,p=0.7, list=FALSE)
validate <- acc_data_train[-inTrain,]

pred <- predict(modFit,validate)
validate$rightpred <- pred == validate$classe

##plots
qplot(roll_belt,yaw_belt,colour=rightpred,data=validate,main="Validation Set Predictions")
```

## 4. Testing Model 
Testing the model built on the given test set

```{r}
pred_test <- predict(modFit,acc_data_test)
```

## 5. Predicting Results

We get the results as :

```{r}
pred_test
```




Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
